---
title: "Simple plotting options for radiocarbon dates in c14_date_lists"
author: "Clemens Schmid"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simple plotting options for radiocarbon dates in c14_date_lists}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

This vignette shows some basic workflows to plot radiocarbon dates in `c14_date_list`s. This is only a short compilation to get you started with c14bazAAR output and [ggplot2](https://ggplot2.tidyverse.org/). 

So let's get started by loading the main packages for this endeavour: c14bazAAR and ggplot2. We will use functions from other packages as well, but address them individually with the `::` operator.

```{r}
library(c14bazAAR)
library(ggplot2)
```

The basis for this example code is the [aDRAC data collection](https://github.com/dirkseidensticker/aDRAC) by Dirk Seidensticker. So let's download the data with the relevant c14bazAAR getter function: 

```{r}
aDRAC <- get_aDRAC()
```

## Temporal plotting of radiocarbon ages

Radiocarbon dating is a method to determine the abolute age of samples. Therefore one of the main aims for plotting naturally is to display temporal information. Let's select the dates of one individual site -- *Batalimo* in Central Africa -- as a subset to reproduce two of the most common types of radiocarbon date plots.

```{r}
Batalimo <- aDRAC %>%
  dplyr::filter(site == "Batalimo")
```

If age modelling and date plotting on a local or regional scale is the major aim of your analysis, you might want to take a look at the [oxcAAR](https://CRAN.R-project.org/package=oxcAAR) package. It serves as an R interface to the [OxCal](https://c14.arch.ox.ac.uk/oxcal.html) software and provides powerful default plotting options.

### Ridgeplots of density distributions

One way to plot radiocarbon dates is to show the probability density distribution of individual calibrated dates as ridgeplots. To produce a plot like that, we need the age-probability information for each plot. We can calculate that with the function `c14bazAAR::calibrate()`. 

```{r, include = FALSE}
Batalimo_calibrated <- Batalimo %>%
  calibrate(choices = "calprobdistr")
```

```{r, eval = FALSE}
Batalimo_calibrated <- Batalimo %>%
  calibrate(choices = "calprobdistr")
```

This adds a list column `calprobdistr` to the input `c14_date_list`. The list column contains a nested data.frame for each date with its probability distribution. 

```{r, echo=FALSE}
Batalimo_calibrated
```

With `tidyr::unnest()` the list column can be dissolved and integrated into the initial `c14_date_list`. Of course the latter looses its original structure and meaning with this step. Each row now represents the probability for one date and year.

```{r}
Batalimo_cal_dens <- Batalimo_calibrated %>% tidyr::unnest()
```

A table like that can be used for plotting a ridgeplot. 

```{r}
Batalimo_cal_dens %>%
  ggplot() +
  # a special geom for ridgeplots is provided by the ggridges package
  ggridges::geom_ridgeline(
    # the relevant variables that have to be mapped for this geom are 
    # x (the time -- here the calibrated age transformed to calBC), 
    # y (the individual lab number of the dates) and
    # height (the probability for each year and date) 
    aes(x = -calage + 1950, y = labnr, height = density),
    # ridgeplots lack a scientifically clear y axis for each 
    # distribution plot and we can adjust the scaling to our needs
    scale = 300
  ) +
  xlab("age calBC/calAD") +
  ylab("dates")
```

### Calcurve plot

Another way to plot radiocarbon dates is to project them onto a calibration curve. The [Bchron](https://CRAN.R-project.org/package=Bchron) R package contains a data.frame with the [intcal13](https://www.doi.org/10.2458/azu_js_rc.55.16947) calibration curve data. We can load the `intcal13` table directly from Bchron.

```{r}
load(system.file('data/intcal13.rda', package = 'Bchron'))
```

For this kind of plot it is more convinient to work with the simplified `calrange` output of `c14bazAAR::calibrate()`.

```{r, include = FALSE}
Batalimo_calibrated <- Batalimo %>%
  calibrate(choices = "calrange")
```

```{r, eval = FALSE}
Batalimo_calibrated <- Batalimo %>%
  calibrate(choices = "calrange")
```

Like the `calprobdistr` option this also adds a list column to the input `c14_date_list`, but a much shorter one for each date only with the age ranges that make up the 2-sigma significance interval of the probability distribution.

```{r}
Batalimo_calibrated$calrange[1:3]
```

Thre resulting table can also be unnested to make the list column content available in the main table.

```{r}
Batalimo_cal_range <- Batalimo_calibrated %>% tidyr::unnest()
```

Now we can plot the calibration curve and -- on top -- error bars with the `calrange` sequences.

```{r, warning=FALSE}
ggplot() +
  # line plot of the intcal curve
  geom_line(
    data = intcal13,
    # again we transform the age information from BP to BC
    mapping = aes(x = -V1 + 1950, y = -V2 + 1950)
  ) +
  # the errorbars are plotted on top of the curve
  geom_errorbarh(
    data = Batalimo_cal_range,
    mapping = aes(y = -c14age + 1950, xmin = -to + 1950, xmax = -from + 1950)
  ) +
  # we define the age range manually -- typically the calcurve
  # is arranged to go from the top left to the bottom right corner
  xlim(-1000, 2000) +
  ylim(2000, -1000) +
  xlab("age calBC/calAD") +
  ylab("uncalibrated age BC/AD")
```

## Spatial mapping of radiocarbon dates

Most radiocarbon dates that can be accessed with c14bazAAR have coordinate information for the respective sites where the samples were taken. Spatial maps therefore are an important form of data visualisation as well.

`c14_date_list`s can directly be transformed to objects of class `sf`. `sf` were introduced by the R package [sf](https://CRAN.R-project.org/package=sf) which provides a very modern and tidy interface to work with spatial data in R.

```{r}
aDRAC_sf <- aDRAC %>% as.sf()
```

This tabular data structure contains the spatial point information for each date in a column *geom*, but also the initial columns of the input dataset: *data.\**

```{r, echo=FALSE}
aDRAC_sf %>% dplyr::select(data.labnr, data.c14age, data.c14std, geom)
```

It can be manipulated with the powerful dplyr functions. We `filter` out all dates from one particular puplication (*Moga 2008*), `group` the dates `by` *site* and apply the `summarise` command to keep only one value per group. As we do not define an operation to fold the other variables in the input table they are removed. Only the geometry column remains.

```{r}  
Moga_spatial <- aDRAC_sf %>%
  dplyr::filter(grepl("Moga 2008", data.shortref)) %>%
  dplyr::group_by(data.site) %>%
  dplyr::summarise()
```

### Interactive map view

The resultuting `sf` object can be plotted interactively with the [mapview](https://CRAN.R-project.org/package=mapview) package. 

```{r}
Moga_spatial %>% mapview::mapview()
```

### Static map plot

The `sf` object can also be used for a static plot -- useful for publications. We download some simple country border base map vector data with the [rnaturalearth](https://CRAN.R-project.org/package=rnaturalearth) R package and transform it to `sf` as well.

```{r}
countries <- rnaturalearth::ne_countries() %>% sf::st_as_sf()
```

Now we can combine the base layer and our point data to create the prototype of a static map plot. 

```{r, warning=FALSE}
ggplot() +
  # geom_sf is a special geom to handle spatial data in sf format
  # the explicit mapping of variables is not necessary here, as geom_sf 
  # automcatically selects the *geom* column in the input table
  geom_sf(data = countries) +
  geom_sf_text(data = countries, mapping = aes(label = formal_en), size = 2) +
  geom_sf(data = Moga_spatial) +
  # with geom_sf comes coord_sf to manage the underlying coordinate grid
  coord_sf(xlim = c(10, 30), ylim = c(0, 15))
```
